---
title: "ENVS-193DS_spring-2025_final"
author: "Ethan Mathews"
date: "06/00/2025"
format:
  html:
    toc: true 
---

link to GitHub repository: https://github.com/ethan-mathews24/ENVS-193DS_spring-2025_final


```{r}
#| message: false

# Reading in Packages

library(tidyverse)

library(gt)

library(janitor)

library(here)

library(readxl)

library(dplyr)

library(tidyr)

library(fs) # fs::dir_tree(path = ".", recurse = TRUE) in console to get file structure 

sst <- read.csv(here("data", "SST_update.csv"))


```


## Problem 1. Research writing


#### a. Transparent statistical methods 




#### b. More information needed




#### c. Suggestions for rewriting 




## Problem 2. Data Visualization 


#### a. Cleaning and summarizing 

```{r}

sst_clean <- sst |> # starting dataframe
  select(date, temp) |>  # selecting only the necessary columns 
  mutate( # creating new columns using the mutate function
    year = year(date), # new column name that is populated with the year by using lubridate package to pull only the year from the date column
    month = month(date, #  new column name that is populated with the month by using the lubridate package 
                  label = TRUE, #returning the month as a factor instead of a numerical value
                  abbr = TRUE) # returning the abbreviated month name instead of the full name
  ) |> 
  filter(year %in% c(2018, 2019, 2020, 2021, 2022, 2023)) |>   # using the filter function to only select the desired years that we want to plot, also reducing processing load 
  select(-date) |>  # removing the original date column 
  group_by(year, month) |> # grouping by year and month 
  summarize(mean_monthly_sst = mean(temp, na.rm = TRUE), # calculating the mean of the surface temp
            .groups = "drop") |>  
  mutate( 
    mean_monthly_sst = round(mean_monthly_sst, 1), # rounding the mean_monthly_sst to one decimal place
    year = as.factor(year)) # making sure the years are seen as a factor
  
```

```{r}

sst_clean |> 
  slice_sample(n = 5)

```

```{r}

str(sst_clean)

```


#### b. Visualize the data 

```{r}

ggplot(data = sst_clean, # using cleaned dataframe
       aes(x = month, # assigning the x-axis
           y = mean_monthly_sst, # assigning the y-axis
           color = year, # coloring by years
           group = year)) + # grouping points by years 
  
  # first layer: line
  geom_line(linewidth = 0.5) + # customizing the line width 
  
  # seoond layer: points
  geom_point(size = 1.1) + # customizing the point size
  
  scale_color_manual(values = c(
      "2018" = "#add8e6",  
      "2019" = "#87ceeb",  
      "2020" = "#4682b4",  
      "2021" = "#4169e1",  
      "2022" = "#0000cd",  
      "2023" = "#00008b")) + # manually adding colors to the different years
  
  labs(x = "Month", # changing the x-axis title 
      y = "Mean monthly sea surface temperature (°C)", # changing the y-axis title 
      color = "Year") + # changing the title of the legend 
  
  theme_bw() + # base theme to add a line to the border 
  
  scale_y_continuous(limits = c(NA, 20)) + # making 20 to be the upper limit of the y-axis 
  
  theme(panel.background = element_rect(fill = "white"), # changing the background color
  panel.grid.major = element_line(color = NA), # removing the color to the grid lines
  panel.grid.minor = element_line(color = NA), # removing the color to the grid lines
  axis.title.x = element_text(size = 9),  # changing size of x axis title
  axis.title.y = element_text(size = 9), # changing size of y axis title
  legend.position = "inside", # moving the legend to be inside of the figure 
  legend.position.inside = c(0.1, 0.75) # changing the position of the legend to be inside the graph by treating it as (x,y) coordinates 
)
  

```



## Problem 3. Data Analysis




#### a. Response variable 




#### b. Purpose of study 




#### c. Difference in "seasons"




#### d. Table of models 




#### e. Run the models 




#### f. Check the diagnostics




#### g. Select the best model 




#### h. Visualize the model predictions 




#### i. Write a caption for your figure 




#### j. Calculate model predictions 




#### k. Interpret the results 




## Problem 4. Affective and exploratory visualizations


#### a. Comparing visualizations 



#### b. Sharing your affective visualization 










